How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2021-04-17 2.04-BigIdea.Rmd
Add files via upload

## What's the big idea? {#bigidea}
In the previous chapters, I've talked about the importance of having a hypothesis, and building that hypothesis in a logical framework within the introduction (see part 3). The introduction serves to inform the reader about why this particular hypothesis was chosen, introducing both the response and determinate variables, as well as the presumed mechanism by which the hypothesis can be falsified (or upheld).

### The funnel introduction 
But so much of what you read doesn't have any clear hypothesis, and there doesn't seem to be any thought about building on bigger concepts or ideas. In this chapter, I suggest that this might not be true only of papers on herpetological invasions (my specialist area), but also of herpetology in general (although I concede that some areas, such as herpetological physiology are actually quite good). Further, I contend that using these wider hypotheses or theories would actually be good for the authors concerned, as it would likely garner them a wider audience. Moreover, a greater number of biologists might come to realise how valuable reptiles and amphibians are as models in biology.

@@

### So where would we find all of these big ideas?
There are quite a few papers that synthesise hypotheses in various areas of biology. Here I provide two to get you started. Ask your advisor about relevant reviews of hypotheses in your area of biological sciences. The first is by Mark Velland [-@vellend2010conceptual] on theories in community ecology. The next is by Jane Catford and colleagues [-@catford2009reducing] on hypotheses in invasion biology, but I encourage you to look for more up to date versions. Each of these papers will give you a list of big ideas, together with the citations for seminal papers that have built them. You will note that many of these theories are very old with some dating back to Darwin. You can think of the way that such ideas are structured as a hierarchy of hypotheses (Figure \@ref(fig:JeschkeHagar61)).

(ref:JeschkeHagar61) **A hierarchy of hypothses can be used to help determine your Big Idea.** A useful structure for thinking about how hypotheses are structured was presented by @heger2018hierarchy in what they termed the _Hierarchy of Hypotheses_

>>>>

# What's the big idea? {#bigidea}
In the previous chapters, I've talked about the importance of having a hypothesis, and building that hypothesis in a logical framework within the introduction (see part 3). The introduction serves to inform the reader about why this particular hypothesis was chosen, introducing both the response and determinate variables, as well as the presumed mechanism by which the hypothesis can be falsified (or upheld).

## The funnel introduction 
But so much of what you read doesn't have any clear hypothesis, and there doesn't seem to be any thought about building on bigger concepts or ideas. In this chapter, I suggest that this might not be true only of papers on herpetological invasions (my specialist area), but also of herpetology in general (although I concede that some areas, such as herpetological physiology are actually quite good). Further, I contend that using these wider hypotheses or theories would actually be good for the authors concerned, as it would likely garner them a wider audience. Moreover, a greater number of biologists might come to realise how valuable reptiles and amphibians are as models in biology.

@@

## So where would we find these big ideas?
There are quite a few papers that synthesise hypotheses in various areas of biology. Here I provide a few to get you started. Ask your advisor about relevant reviews of hypotheses in your area of biological sciences. The first is by Mark Velland [-@vellend2010conceptual] on theories in community ecology. The next is by Jane Catford and colleagues [-@catford2009reducing] on hypotheses in invasion biology, but I encourage you to look for more up to date versions. Lastly, consider these wide ranging papers about so-called 'laws' in biology and ecology [@dhar2010laws; @lawton1999are]. Each of these papers will give you a list of big ideas, together with the citations for seminal papers that have built them. You will note that many of these theories are very old with some dating back to Darwin. You can think of the way that such ideas are structured as a hierarchy of hypotheses (Figure \@ref(fig:JeschkeHagar61)). Building such a hierarchy of ideas in your own specialist subject area is good reference material to prompt [critical reading](critical.html) and reflection. You may want to take this structure forward when thinking how to [introduce your PhD chapters](PhDreview.html).

(ref:JeschkeHagar61) **A hierarchy of hypothses can be used to help determine your Big Idea.** A useful structure for thinking about how hypotheses are structured was presented by @heger2018hierarchy in what they termed the _Hierarchy of Hypotheses_.

### 2021-04-17 2.06-argument.Rmd
Add files via upload

## Construct a logical argument in your writing {#argument}

@@

### Arguments in Latin
From time to time you may come across comments about arguments that you’ve made in your work made in Latin. Why do people use Latin and not simply say it in English? What do these phrases actually mean? While you may think that these people are being pompous or trying to demonstrate some scholarly superiority by quoting Latin at you (and you might be right), there is some important reasoning to consider both in the meanings of these argumentum (as they are known in Latin), and the fact that they are as old as Roman scholars, and often much older. This then is the point, that some arguments are faulty, and that this has been known for thousands of years.

@@

_Novacula Occami_ - "Occam’s razor": The idea that the simplest hypothesis, or argument is most likely to be correct. Another way to think of this is if you have two models that you test are equally likely (e.g. within 2 dAICs), the one with the least number of variables should be considered superior. This is also known as the priciple or law of parsimony: _lex parsimoniae_.

>>>>

# Construct a logical argument in your writing {#argument}

@@

## Arguments in Latin
From time to time you may come across comments about arguments that you’ve made in your work made in Latin. Why do people use Latin and not simply say it in English? What do these phrases actually mean? While you may think that these people are being pompous or trying to demonstrate some scholarly superiority by quoting Latin at you (and you might be right), there is some important reasoning to consider both in the meanings of these argumentum (as they are known in Latin), and the fact that they are as old as Roman scholars, and often much older. This then is the point, that some arguments are faulty, and that this has been known for thousands of years.

@@

_Novacula Occami_ - "Occam’s razor": The idea that the simplest hypothesis, or argument is most likely to be correct. Another way to think of this is if you have two models that you test are equally likely (e.g. within 2 δAICs), the one with the least number of variables should be considered superior. This is also known as the priciple or law of parsimony: _lex parsimoniae_.

## Other arguments in English

### 2021-04-17 2.20-proposal.Rmd
Add files via upload

## What is needed for your research proposal? {#proposal}

@@

### Structure

We expect that your final thesis will be a set of papers (data chapters sandwiched between an introduction and discussion), and so we ask you to prepare the proposal in the same way. That is, you will normally have two (MSc) or five (PhD) data chapters. In the proposal, each chapter will consist of three parts:

### Proposal introduction

@@

### Methods and Materials

@@

### Hypothetical Results

@@

### Your Proposal - what is it good for?

@@

#### Copy and paste directly into your thesis

@@

#### Use it to raise money to do your studies.

@@

#### Deposit the proposal to conform to Design and Analysis Transparency

>>>>

# What is needed for your research proposal? {#proposal}

@@

## Structure

We expect that your final thesis will be a set of papers (data chapters sandwiched between an [introduction and discussion](PhDintro.html)), and so we ask you to prepare the proposal in the same way. That is, you will normally have two (MSc) or five (PhD) data chapters. In the proposal, each chapter will consist of three parts:

## Proposal introduction

@@

## Methods and Materials

@@

## Hypothetical Results

@@

## Your Proposal - what is it good for?

@@

### Copy and paste directly into your thesis

@@

### Use it to raise money to do your studies.

@@

### Deposit the proposal to conform to Design and Analysis Transparency

### 2021-04-17 3.01-Title.Rmd
Add files via upload

# Writing the sections that make up your chapter  {#part3}

@@

### The Title

@@

#### Some title ideas to start you off
- Don’t start by looking for the best title
  + Write a number of candidate titles and ask your co-authors to vote for their favorites.
- The shorter and catchier your title can be the better: 6 words (±1) is an ideal.
  + Allow yourself a longer subtitle if needed, but don’t go over 20 words total (some journals may limit your total to less). 
  + Consider the (former) 120 character limit of a Tweet as an upper limit
- Do include your principal finding if possible
- Include as many key words as you can

#### Things to avoid in your title
 - Don’t feel obliged to include [taxonomic terms](taxonomy.html) unless it is relevant or compulsory
   + Some (taxonomically minded) journals will insist on the species name followed by the taxonomic authority, and/or the family and order, in your title
 - Avoid obscure or specialist words that won’t be understood by your readership
   + There are times when key words are necessarily specialised and your readership will expect this, but simpler words in your title will open up your readership which will otherwise remain narrow
 - Don’t simply define the scope of your work without including your content.
 - Don’t have your title as a question. Rather provide the answer!

### Key Words
The key words are a way for readers to find your content with searches. Typically, the advice is to use words that are not in the title or abstract. This is because many databases have combined searching facilities for title, abstract and key words. I struggle to think of appropriate key words. Some of the big idea words and short phrases from the introduction. Then I tend to look at articles in the same genre, and see what key words they have used. As with the title, I suggest that you enter your chosen key words into your literature database of choice [see part 2](database.html) and see what comes back. You should see a group of papers that look wholly familiar and preferably those that are already cited in your manuscript. If not, it’s time to look again.

>>>>

# Writing the sections that make up your data chapter  {#part3}

@@

## The Title

@@

### Some title ideas to start you off
- Don’t start by looking for the best title
  + Write a number of candidate titles and ask your co-authors to vote for their favorites.
- The shorter and catchier your title can be the better: 6 words (±1) is an ideal.
  + Allow yourself a longer subtitle if needed, but don’t go over 20 words total (some journals may limit your total to less). 
  + Consider the (former) 120 character limit of a Tweet as an upper limit
- Do include your principal finding if possible
- Include as many key words as you can

### Things to avoid in your title
 - Don’t feel obliged to include [taxonomic terms](taxonomy.html) unless it is relevant or compulsory
   + Some (taxonomically minded) journals will insist on the species name followed by the taxonomic authority, and/or the family and order, in your title
 - Avoid obscure or specialist words that won’t be understood by your readership
   + There are times when key words are necessarily specialised and your readership will expect this, but simpler words in your title will open up your readership which will otherwise remain narrow
 - Don’t simply define the scope of your work without including your content.
 - Don’t have your title as a question. Rather provide the answer!

## Key Words
The key words are a way for readers to find your content with searches. Typically, the advice is to use words that are not in the title or abstract. This is because many databases have combined searching facilities for title, abstract and key words. I struggle to think of appropriate key words. Some of the big idea words and short phrases from the introduction. Then I tend to look at articles in the same genre, and see what key words they have used. As with the title, I suggest that you enter your chosen key words into your literature database of choice [see part 2](database.html) and see what comes back. You should see a group of papers that look wholly familiar and preferably those that are already cited in your manuscript. If not, it’s time to look again.

### 2021-04-17 3.05-Results.Rmd
Add files via upload

## The Results {#results}

@@

### The correct approach - responding to your hypotheses
Yes, the easiest way to write the results is simply to have the questions or hypotheses that you posed at the end of the introduction and have them as subheadings for your results section. Once you have these subheadings, you simply assemble all of the results with respect to that hypothesis. You should already have all the tests that you used in the methods section.

@@

If you do have to insert the results of a statistical test into the text, include the test statistic, degrees of freedom (which incorporates N - the sample size) and the value of P. Only if P is less than alpha (usually 0.05) can you use the word ‘significant’ when describing the effect.

@@

### When to have a table or graph?
Most journals will allow you to have as many tables or graphs as you need to give your results. There are a few important things to consider. First, that data shouldn’t be repeated. So don’t have a table of results where the same results are also in a graph (or even two graphs that show the same data). Also, don’t repeat data from a figure or table in your text. You need to decide how best to present your data. If you can’t decide then discuss it with your advisor by presenting them with the alternatives. Good results should convey your findings clearly, and consulting with others should quickly help you decide which approach is effective.

@@

### Descriptive results
Some studies have a set of descriptive results that, by convention, are placed without a section header at the start of the results. These normally convey the size of the dataset that entered into the rest of the analyses. If the size of the dataset varies by design in the methods, then you don’t need it here. But you can’t always control the sample size, so it needs to be established before moving on to the rest of the results. In genetic studies, for example, this might include the number of samples that were collected and successfully sequenced, and thus entered into the rest of the analyses. Your text can then be broken into sections that correspond with your questions, unless you only have one in which case further sections are not warranted.

### Responding to hypotheses
While I encourage you to present data that corresponds to hypotheses here in the results section, you should not formally accept or reject your hypothesis here. This is done in the discussion, and allows you to present caveats and other important issues.

### Common Mistakes
- Having extra analyses in the results that you didn’t include in your methods. If you really didn’t anticipate conducting the extra tests, then these are actually _post hoc_ (after the event) tests (not anticipated before the study) and you should think carefully about how to treat these. Otherwise, you’ll need to r-ejig your methods to make sure that all tests are included.
- Meaningless precision. This can happen throughout a thesis or manuscript (e.g. in the Materials and Methods), but it is most common in the results when too many significant figures are reported. For example, the output of your analysis might tell you that flight feathers in one population are 2.6484332 mm longer than those of another (± 0.0654562 mm), but in your methods you only use callipers that have a precision of 0.01 mm. Reporting results beyond the precision whereby they can be measured quickly becomes meaningless. Whenever you obtain the output of a statistical package, you need to make sure that you use your practical head to re-interpret these results and think about what they actually mean. 
- Repeating numbers in the text that are already in tables or graphs. The text is likely to be packed with numbers already, so don’t repeat any that are already represented elsewhere.
- Repeating text in different sections. As you’ve seen elsewhere in this book, repeating text simply isn’t acceptable. It doesn’t matter if you did the same thing many times, you can always write something in a different way. Having repeated text is the fastest way to bore the pants off your reader, and therefore must be avoided at all costs.

>>>>

# The Results {#results}

@@

## The correct approach - responding to your hypotheses
Yes, the easiest way to write the results is simply to have the questions or hypotheses that you posed at the end of the introduction and have them as subheadings for your results section. Once you have these subheadings, you simply assemble all of the results with respect to that hypothesis. You should already have all the tests that you used in the methods section.

@@

If you do have to insert the results of a statistical test into the text, include the test statistic, degrees of freedom (which incorporates N - the sample size) and the value of P. Only if P is less than α (alpha is usually 0.05) can you use the word ‘significant’ when describing the effect.

@@

## When to have a table or graph?
Most journals will allow you to have as many tables or graphs as you need to give your results. There are a few important things to consider. First, that data shouldn’t be repeated. So don’t have a table of results where the same results are also in a graph (or even two graphs that show the same data). Also, don’t repeat data from a figure or table in your text. You need to decide how best to present your data. If you can’t decide then discuss it with your advisor by presenting them with the alternatives. Good results should convey your findings clearly, and consulting with others should quickly help you decide which approach is effective.

@@

## Descriptive results
Some studies have a set of descriptive results that, by convention, are placed without a section header at the start of the results. These normally convey the size of the dataset that entered into the rest of the analyses. If the size of the dataset varies by design in the methods, then you don’t need it here. But you can’t always control the sample size, so it needs to be established before moving on to the rest of the results. In genetic studies, for example, this might include the number of samples that were collected and successfully sequenced, and thus entered into the rest of the analyses. Your text can then be broken into sections that correspond with your questions, unless you only have one in which case further sections are not warranted.

## Responding to hypotheses
While I encourage you to present data that corresponds to hypotheses here in the results section, you should not formally accept or reject your hypothesis here. This is done in the discussion, and allows you to present caveats and other important issues.

## Common Mistakes
- Having extra analyses in the results that you didn’t include in your methods. If you really didn’t anticipate conducting the extra tests, then these are actually _post hoc_ (after the event) tests (not anticipated before the study) and you should think carefully about how to treat these. Otherwise, you’ll need to r-ejig your methods to make sure that all tests are included.
- Meaningless precision. This can happen throughout a thesis or manuscript (e.g. in the Materials and Methods), but it is most common in the results when too many significant figures are reported. For example, the output of your analysis might tell you that flight feathers in one population are 2.6484332 mm longer than those of another (± 0.0654562 mm), but in your methods you only use callipers that have a precision of 0.01 mm. Reporting results beyond the precision whereby they can be measured quickly becomes meaningless. Whenever you obtain the output of a statistical package, you need to make sure that you use your practical head to re-interpret these results and think about what they actually mean. 
- Repeating numbers in the text that are already in tables or graphs. The text is likely to be packed with numbers already, so don’t repeat any that are already represented elsewhere.
- Repeating text in different sections. As you’ve seen elsewhere in this book, repeating text simply isn’t acceptable. It doesn’t matter if you did the same thing many times, you can always write something in a different way. Having repeated text is the fastest way to bore the pants off your reader, and therefore must be avoided at all costs.

### 2021-04-17 index.Rmd
Add files via upload

--- 
title: "How to write a PhD in Biological Sciences: a guide for the uninitiated"
author: "John Measey"
date: "`r Sys.Date()`"
documentclass: krantz
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
colorlinks: yes
lot: yes
lof: yes
site: bookdown::bookdown_site
description: "You don’t have to be a genius to write a PhD. It will always involve a lot of hard work and dedication, but the process of writing up is a whole lot easier if you understand the basic ground rules. This book is a guide specifically for writing a PhD in the *biological sciences*. It will guide you from deciding whether or not you should actually do a PhD, provide practical, up-to-the-minute guides to getting started with planning and writing, and consolidate this with the nuts and bolts of writing for the biological sciences. The book guides you through the do’s and don’ts of writing a thesis, and taking your thesis chapters to publication in peer-reviewed journals. Whether you are a student just starting your studies, or a supervisor struggling to cope, this book will provide the insider information you need to get ahead."
github-repo: yihui/bookdown-crc
graphics: yes
#cover-image: images/cover.png
---

@@

**The book is divided into three main parts:**

@@

[**Part 2**](part2.html)  
Part two of the book sets out a practical guide to getting started with writing. The approach I've taken here is describing the production of a PhD in biological sciences which is made up of data chapters, and where each chapter is essentially a potential publication in a science journal. If you didn't want to do a PhD but just wanted to write a publication, then most of what is here (and in Part 3) will be of use as you could take the same formulaic approach as given in this book. Similarly, if you were writing a MSc project, this section would also be useful. There is a lot of technical stuff involved in writing science, and in part 2 I cover as much as I can while still remaining general to the subject area.

@@

@@

## Software information and conventions {-}

I used the **knitr**\index{knitr} package [@xie2015dynamic] and the **bookdown**\index{bookdown} package [@R-bookdown] to compile my book.

>>>>

--- 
title: 'How to write a PhD in Biological Sciences: a guide for the uninitiated'
author: "John Measey"
date: "`r Sys.Date()`"
documentclass: krantz
bibliography:
- book.bib
- packages.bib
biblio-style: apalike
link-citations: yes
colorlinks: yes
lot: yes
lof: yes
site: bookdown::bookdown_site
description: You don’t have to be a genius to write a PhD. It will always involve
  a lot of hard work and dedication, but the process of writing up is a whole lot
  easier if you understand the basic ground rules. This book is a guide specifically
  for writing a PhD in the *biological sciences*. It will guide you from deciding
  whether or not you should actually do a PhD, provide practical, up-to-the-minute
  guides to getting started with planning and writing, and consolidate this with the
  nuts and bolts of writing for the biological sciences. The book guides you through
  the do’s and don’ts of writing a thesis, and taking your thesis chapters to publication
  in peer-reviewed journals. Whether you are a student just starting your studies,
  or a supervisor struggling to cope, this book will provide the insider information
  you need to get ahead.
github-repo: yihui/bookdown-crc
graphics: yes
---

@@

**The book is divided into four main parts:**

@@

[**Part 2**](part2.html)
Part two of the book sets out a practical guide to getting started with writing. The approach I've taken here is describing the production of a PhD in biological sciences which is made up of data chapters, and where each chapter is essentially a potential publication in a science journal. If you didn't want to do a PhD but just wanted to write a publication, then most of what is here (and in [Part 3](part3.html)) will be of use as you could take the same formulaic approach as given in this book. Similarly, if you were writing a MSc project, this section would also be useful. There is a lot of technical stuff involved in writing science, and in part 2 I cover as much as I can while still remaining general to the subject area.

@@

[**Part 4**](part4.html)
The concluding part of this book is written to cover the things that remain after you have finished writing all of your data chapters. Most importantly, these include the introductory and concluding chapters that sit as bookends to the data chapters that you have already produced. These chapters make your thesis into a 'body of work', and allow you space to reflect on what you have achieved in your PhD studies.

@@
